A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-30 data.tex
Sending bits and pieces of text to the workhouse

A reproduction package featuring all relevant logs and all benchmarked instances is available~\cite{artefact}.
\Catra{} and \Ostrich{} are both available as living software
under a 3-clause BSD~license on GitHub at \url{https://github.com/amandasystems/catra} and \url{https://github.com/uuverifiers/ostrich}
respectively. A version of \Ostrich{} modified to use CA-Str is available in the \texttt{cea-catra} branch. All source code to generate the paper itself
is available at \url{https://github.com/amandasystems/oopsla-artefact}.

>>>>

A reproduction package featuring all relevant logs and all benchmarked instances is available~\cite{artefact}.
\Catra{} and \Ostrich{} are both available as living software
under a 3-clause BSD~license on GitHub at \url{https://github.com/amandasystems/catra} and \url{https://github.com/uuverifiers/ostrich}
respectively. A version of \Ostrich{} using CA-Str is available in the \texttt{cea-catra} branch. Code for the paper itself
is available at \url{https://github.com/amandasystems/oopsla-artefact}.

### 2024-03-30 multiple.tex
Bundle a new version of acm template
Sending bits and pieces of text to the workhouse

      % EXPAND
      \ExpandM{} & 
      \inferrule%
      {
       {\begin{array}{c@{}}
           \left\{
            \begin{array}{c@{}}
              \FlowEq(\Automaton_i, \Filter),
              \Connected(\Automaton_i, \Filter),\\[1ex]
              \MonoidElement = \sum_{\Transition \in \Transitions_{\Automaton_i}} \Filter(\Transition) \cdot \Map(\Transition)
            \end{array}
            \right\}_{i=1}^k,
           \\[4ex]
          \ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, ~
          \SomeInequalities, \SomeClause
        \end{array}}
        }
      {\ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} & 
      None \\[5ex]
      \Materialise{} &
      \inferrule%
      {{
         \begin{array}{c@{}}
           \FlowEq(\Automaton', \Filter'),
           \Connected(\Automaton', \Filter'),
           \BindingSum(\Automaton_i, \Automaton_j, \Automaton', \Filter'),\\
           \ImagePredicate{
                     \Tuple{%
                      \Automaton_1,  \Automaton_{i-1},  \Automaton_{i+1},
                      \ldots,  \Automaton_{j-1},  \Automaton_{j+1},
                      \ldots,  \Automaton_k,  \Automaton'
                     }
           }{\Map}{\Filter'}{\MonoidElement},
\SomeInequalities, \SomeClause
           \end{array}}
           }
      {{
      \begin{array}{l@{}}
      \ImagePredicate{
        \Tuple{%
          \Automaton_1,  \ldots,  \Automaton_i, \ldots, 
          \Automaton_j, \ldots,  \Automaton_k
        }}{\Map}{\Filter}{\MonoidElement}
        \SomeInequalities, \SomeClause{}
      \end{array}}} &%
      $\begin{array}{l@{}}
          \Filter' = \Extend(\Automaton',\Filter),\\
          1 \leq i < j \leq k, \\
       \Automaton' =
         \Filter_{\SomeInequalities}(\Automaton_i) \times \Filter_{\SomeInequalities}(\Automaton_j)\\
      \end{array}$ \\
      \bottomrule
      \end{tabular}
    \end{table}

@@

For instances of precisely one automaton, neither rule applies, and we perform
the calculus as before.

@@

We then continue by repeating the steps of~\cref{sec:single:example}, shown 
in~\cref{fig:derivation:single} since the steps to propagate
\cref{ex:string-constraints} across the automaton $\mathcal{A}'$ still apply for the
product. Note that this fixes the values of all transitions but one of $\mathcal{A}'$ as
constants, including the free counter variables $i, n$, whose new constant
values we propagate everywhere.

Once we have performed reasoning by equality to eliminate transition variables,
we obtain the version of $\mathcal{B}$ shown in \cref{fig:aut_b_annotated} and represented
by the clauses in the final node in the tree. This corresponds to the end of
\cref{sec:intuition:algebra}. Our only options now are to either split on which
path we take from $I_B$ by case-splitting on~$\TransitionVar_{12}$ or to
directly invoke \Materialise{}. The latter would produce a shorter tree in this
instance but might lead to a larger product being computed, so we (and indeed
\Catra{}) pick \Split{}. This leads to an opportunity for propagation since
$\TransitionVar_{12} = 0$ cuts off $\TransitionVar_{9}$ from $I_B$ on the left
branch. Note that the same opportunity does not present itself on the right
branch, though; the use of the upper path to state~$A$ does not preclude visiting
state~$B$.

In both cases, we have comparatively simple automata left, so we apply
\Materialise{} to obtain their product, which for both branches is empty. An
empty product will have no outgoing transitions from its initial state, and so
will produce a flow equation $1 = 0$ since the right-hand balancing side will
always be the empty sum. We can then close the proof and know that the problem
is unsatisfiable. A full derivation tree for the example can be found 
in~\cref{fig:derivation:multi}.

>>>>

      % EXPAND
      \ExpandM{} & 
      \inferrule%
      {
       {\begin{array}{c@{}}
           \left\{
            \begin{array}{c@{}}
              \FlowEq(\Automaton_i, \Filter),
              \Connected(\Automaton_i, \Filter),\\[1ex]
              \MonoidElement = \sum_{\Transition \in \Transitions_{\Automaton_i}} \Filter(\Transition) \cdot \Map(\Transition)
            \end{array}
            \right\}_{i=1}^k,
           \\[4ex]
          \ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, ~
          \SomeInequalities, \SomeClause
        \end{array}}
        }
      {\ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} & 
      None \\[5ex]
      \Materialise{} &
      \inferrule%
      {{
         \begin{array}{c@{}}
           \FlowEq(\Automaton', \Filter'),\\
           \Connected(\Automaton', \Filter'),
           \BindingSum(\Automaton_i, \Automaton_j, \Automaton', \Filter'),\\
           \ImagePredicate{
                     \Tuple{%
                      \Automaton_1,  \Automaton_{i-1},  \Automaton_{i+1},
                      \ldots,  \Automaton_{j-1},  \Automaton_{j+1},
                      \ldots,  \Automaton_k,  \Automaton'
                     }
           }{\Map}{\Filter'}{\MonoidElement},
\SomeInequalities, \SomeClause
           \end{array}}
           }
      {{
      \begin{array}{l@{}}
      \ImagePredicate{
        \Tuple{%
          \Automaton_1,  \ldots,  \Automaton_i, \ldots, 
          \Automaton_j, \ldots,  \Automaton_k
        }}{\Map}{\Filter}{\MonoidElement}
        \SomeInequalities, \SomeClause{}
      \end{array}}} &%
      $\begin{array}{l@{}}
          \Filter' = \Extend(\Automaton',\Filter),\\
          1 \leq i < j \leq k, \\
       \Automaton' =
         \Filter_{\SomeInequalities}(\Automaton_i) \times \Filter_{\SomeInequalities}(\Automaton_j)\\
      \end{array}$ \\
      \bottomrule
      \end{tabular}
    \end{table}

@@

For precisely one automaton, neither rule applies, and we perform
the calculus as before.

@@

We then continue by repeating the steps of~\cref{sec:single:example}, shown 
in~\cref{fig:derivation:single} since the steps to propagate
\cref{ex:string-constraints} across the automaton $\mathcal{A}'$ still apply for the
product. Note that this fixes the values of all transitions but one of $\mathcal{A}'$, including the free counter variables $i, n$, whose new constant
values we propagate everywhere.

Once we have performed reasoning by equality to eliminate transition variables,
we obtain the version of $\mathcal{B}$ shown in \cref{fig:aut_b_annotated} and represented
by the clauses in the final node in the tree. This corresponds to the end of
\cref{sec:intuition:algebra}. Our only options now are to either split on which
path we take from $I_B$ by case-splitting on~$\TransitionVar_{12}$ or to
directly invoke \Materialise{}. The latter would produce a shorter tree in this
instance but might lead to a larger product being computed, so we pick \Split{}. This leads to an opportunity for propagation since
$\TransitionVar_{12} = 0$ cuts off $\TransitionVar_{9}$ from $I_B$ on the left
branch. Note that the same opportunity does not present itself on the right
branch; the use of the upper path to state~$A$ does not preclude visiting
state~$B$.

In both cases the automata are now simpler, so we apply
\Materialise{} to obtain their product, which for both branches is empty. An
empty product will have no outgoing transitions from its initial state, and so
will lead to a flow equation $1 = 0$. We can then close the proof and know that the problem
is unsatisfiable. A full derivation tree for the example can be found 
in~\cref{fig:derivation:multi}.

### 2024-03-30 preliminaries.tex
Add logics to preliminaries
more
Sending bits and pieces of text to the workhouse

\subsection{Languages, Finite-state Automata and their Products}
\label{sec:languages}

@@

\subsection{The Parikh Map and its Image}
Formally, the \textit{Parikh map} over an alphabet $\Alphabet=
\left\{a_1, \ldots, a_k \right\}$ is defined as in \cite{kozen}:
$$
\begin{aligned}
& \ParikhMap: \MapFromTo{\Strings}{\natural^k} \\
& \ParikhMap(s) = \VectorLiteral{\#a_1(s), \#a_2(s), \ldots, \#a_k(s)}
\end{aligned}
$$

@@

\subsection{The Parikh image of a regular language expressed in Presburger arithmetic}
\label{sec:verma}

>>>>

%
%\subsection{Logics}\label{sec:preliminaries:logic}
%

\subsection{Languages, Finite-State Automata and Their Products}\label{sec:languages}

@@

\subsection{The Parikh Map and Its Image}
Formally, the \textit{Parikh map} over an alphabet $\Alphabet=
\left\{a_1, \ldots, a_k \right\}$ is defined as in \cite{kozen}:
$$
\begin{aligned}
& \ParikhMap: \MapFromTo{\Strings}{\natural^k} \\
& \ParikhMap(s) = \VectorLiteral{\#a_1(s), \#a_2(s), \ldots, \#a_k(s)}
\end{aligned}
$$

@@

\subsection{The Parikh Image of a Regular Language Expressed in Presburger Arithmetic}
\label{sec:verma}
